A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-17 tex/body.tex
Fix minor things, add paragraph on implicitly computing mu-terms.  Add citation.

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w_1 \teq \const{C}( x ),\;
w_2 \teq \const{E}( w_3 ),\;
x \not\teq y_2,\;
y_1 \teq \const{C}( y_2 ),\;
y_2 \teq \const{D}( x ),\;
z_1 \teq \const{E}( z_2 ),\;
z_2 \teq \const{C}( z_1 )
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
\Val \ec{w_1} & {}\,=\,{} & \MU {\Varec{w_1}}. & \const{C}( {\Varec{x}} ) &
\Val \ec{y_1} & {}\,=\,{} & \MU {\Varec{y_1}}. & \const{C}( & \MU {\Varec{y_2}}. & \const{D}( {\Varec{x}} ) ) \\
\Val \ec{w_2} & = & \MU {\Varec{w_2}}. & \const{E}( {\Varec{w_3}} ) &
\Val \ec{y_2} & = & \MU {\Varec{y_2}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} \\
\Val \ec{w_3} & = & {\Varec{w_3}} &&
\Val \ec{z_1} & = & \MU {\Varec{z_1}}. & \const{E}( & \MU {\Varec{z_2}}. & \const{C}( {\Varec{z_1}} ) ) \\
\Val \ec{x} & = & {\Varec{x}} &&
\Val \ec{z_2} & = & \MU {\Varec{z_2}}. & \const{C}( & \MU {\Varec{z_1}}. & \const{E}( {\Varec{z_2}} ) )
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{w_3}}$ and ${\Varec{x}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$} is
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( \const{E}( x )),\;
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{E}( \const{C}( x ))
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$, and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to
become a non-model.
% "be inconsistent." Models can't be "inconsistent".
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
After applying the substitution for the third term mentioned above\rem{which third term?},
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\paragraph{Integration into {\relax\XXXL($T$)}.}
The calculus is implemented as a \emph{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a input set of
clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We may interpret such a
partial assignment as a set of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$, and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, at a high level, by the Nelson--Oppen combination
framework, each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables of the
same type; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F$, adds a clause $C$ to $F$, or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above. For each equality added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For the \rn{Split} rule, we add a clause of the form
$\const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_1}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver and not by the theory solver. All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 (along with the
optimization for collapsing selectors) and the \rn{Single} rule are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phase~2 and the \rn{Split} rule are performed only for complete satisfying
assignments $M$, and in that order.\rem{in what order?}

\rem{TODO: * sharing is caring (let's cite it for a change)}
\rem{TODO: ``theory combination more complicated (and Theorem Completeness too weak)''?}
\rem{TODO: representation of $\mu$-terms?}

\paragraph{Optimizations.}
Our implementation uses several optimizations not reflected in the presentation
of the calculus, which closely follow the approach outlined by Barrett et al.\
\cite{barrett-et-al-2007}.%
\rem{What closely follows the approach by Barrett et al.? The optimizations or
the calculus?}
We only briefly mention these optimizations, since
each applies to datatypes as well as codatatypes, and most have been presented
thoroughly in Barrett et al.

@@

For each goal in each theory file, Sledgehammer was invoked to select
256~lemmas as axioms, which were then monomorphized and translated to SMT-LIB
along with the goal \cite{boehme-2012-phd}. The resulting problem was given to
CVC4, running for up to \textbf{TODO: XXX} seconds on the StarExec cluster
\cite{stump-et-al-2014-starexec}. Problems involving no (co)datatypes were
filtered out. Due to the lack of machinery for reconstructing inferences about
(co)datatypes in Isabelle, CVC4 is trusted as an oracle in these experiments.

@@

We introduced a decision procedure for the ground theory of datatypes and
codatatypes and implemented it in the SMT solver CVC4. Our main
contribution has been the support for codatatypes. Both the metatheory and
the implementation rely heavily on $\mu$-terms to represent cyclic values,
a complication specific to codatatypes.
% are not very
% difficult per se, but there are tricky corner cases to take care of if we care
% about completeness (which we do).
%% alliteration
On the practical side, we obtained respectable results on Isabelle benchmarks.

@@

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah. \textbf{TODO FIXME}
Morgan Deters?
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler shared their private
theories with us so we could include them in the benchmarks.
Andrei Popescu helped clarify our thoughts regarding the axiomatization of
codatatypes. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).

>>>>

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w_1 \teq \const{C}( x ),\;
w_2 \teq \const{E}( w_3 ),\;
x \not\teq y_2,\;
y_1 \teq \const{C}( y_2 ),\;
y_2 \teq \const{D}( x ),\;
z_1 \teq \const{E}( z_2 ),\;
z_2 \teq \const{C}( z_1 )
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
\Val \ec{w_1} & {}\,=\,{} & \MU {\Varec{w_1}}. & \const{C}( {\Varec{x}} ) &
\Val \ec{y_1} & {}\,=\,{} & \MU {\Varec{y_1}}. & \const{C}( & \MU {\Varec{y_2}}. & \const{D}( {\Varec{x}} ) ) \\
\Val \ec{w_2} & = & \MU {\Varec{w_2}}. & \const{E}( {\Varec{w_3}} ) &
\Val \ec{y_2} & = & \MU {\Varec{y_2}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} \\
\Val \ec{w_3} & = & {\Varec{w_3}} &&
\Val \ec{z_1} & = & \MU {\Varec{z_1}}. & \const{E}( & \MU {\Varec{z_2}}. & \const{C}( {\Varec{z_1}} ) ) \\
\Val \ec{x} & = & {\Varec{x}} &&
\Val \ec{z_2} & = & \MU {\Varec{z_2}}. & \const{C}( & \MU {\Varec{z_1}}. & \const{E}( {\Varec{z_2}} ) )
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{w_3}}$ and ${\Varec{x}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$} is
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( \const{E}( x )),\;
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{E}( \const{C}( x ))
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$, and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to
become a non-model.
% "be inconsistent." Models can't be "inconsistent".
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcal{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
After applying the substitution for $\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$, we have that
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\paragraph{Integration into {\relax\XXXL($T$)}.}
The calculus is implemented as a \emph{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a input set of
clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We may interpret such a
partial assignment as a set of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$, and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, at a high level, by the Nelson--Oppen combination
framework (its variant used in CVC4 is described in~\cite{jovanovic2011sharing}), 
each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables of the
same type; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F$, adds a clause $C$ to $F$, or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above. For each equality added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For the \rn{Split} rule, we add a clause of the form
$\const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_1}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver and not by the theory solver. All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 along with the
optimization for collapsing selectors are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

Prior to constructing a model for $F$, the theory solver does not explicitly construct $\mu$-terms nor the map $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To recognize when the precondition of \rn{Acyclic} holds,
the procedure considers each datatype equivalence class $\ec{t}$ containing a constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes in the same way.
If while doing so it returns to $\ec{t}$, then it deduces that \rn{Acyclic} applies to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $S$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $S$,
while partitioning $S$ into $S_1, \ldots, S_n$,
where all terms in $S_i$ have the same constructor type for all their subterms of type $\Data \cup \Codata$,
and are equivalent for all subterms of type $\Nondata$.
If after doing so, some $S_i$ contains two terms $u$ and $v$, then it deduces that \rn{Unique} applies to $u$ and $v$.

%\rem{TODO: * sharing is caring (let's cite it for a change)}
\rem{TODO: ``theory combination more complicated (and Theorem Completeness too weak)''?}
%\rem{TODO: representation of $\mu$-terms?}

\paragraph{Optimizations.}
Our implementation uses several optimizations which
follow the approach outlined by Barrett et al.\cite{barrett-et-al-2007}
and are not reflected in the presentation of the calculus.
We only briefly mention these optimizations, since
each applies to datatypes as well as codatatypes, and most have been presented
thoroughly in Barrett et al.

@@

For each goal in each theory file, Sledgehammer was invoked to select
256~lemmas as axioms, which were then monomorphized and translated to SMT-LIB
along with the goal \cite{boehme-2012-phd}. The resulting problem was given to
CVC4, running for up to 60 seconds on the StarExec cluster
\cite{stump-et-al-2014-starexec}. Problems involving no (co)datatypes were
filtered out. Due to the lack of machinery for reconstructing inferences about
(co)datatypes in Isabelle, CVC4 is trusted as an oracle in these experiments.

@@

We introduced a decision procedure for the ground theory of datatypes and
codatatypes and implemented it in the SMT solver CVC4. Our main
contribution has been the support for codatatypes. Both the metatheory and
the implementation rely heavily on $\mu$-terms to represent cyclic values,
a complication specific to codatatypes.
% are not very
% difficult per se, but there are tricky corner cases to take care of if we care
% about completeness (which we do).
%% alliteration
On the practical side, we obtained respectable results on Isabelle benchmarks. \rem{emphasize new capabilities allows us to improve state of the art.}

@@

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to the development team of CVC4, 
including Clark Barrett and Cesare Tinelli,
and in particular Morgan Deters,
who jointly the first author developed the theory solver for datatypes in CVC4.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler shared their private
theories with us so we could include them in the benchmarks.
Andrei Popescu helped clarify our thoughts regarding the axiomatization of
codatatypes. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
